\section{Identifiability and cell-total consistency}
\label{sec:identifiability}

We port \cref{thm:bg-id} to ST and derive a cell-total consistency
corollary.

\subsection{The rank condition}

Let the spot-by-cell-type composition matrix $P \in [0,1]^{S \times K}$
have entries $P_{sk} = p_{s,k}$, with rows summing to $1$. For each
gene $j$, the spot-level mean expression is $\E[X_{sj}/N_s] = \sum_k
P_{sk} \mu_{j,k}$. In matrix form, $\bm{m}_j = P \bm{\mu}_j$ where
$\bm{m}_j \in \R^S$ is the vector of per-spot mean expressions and
$\bm{\mu}_j \in \R^K$ is the per-cell-type mean expression vector
for gene $j$.

\begin{theorem}[Identifiability of cell-type-specific expression]
\label{thm:rank}
Assume Poisson sampling $X_{sj} \sim
\mathrm{Poisson}\bigl(N_s \sum_k p_{s,k} \mu_{j,k}\bigr)$ with the
composition matrix $P$ known (e.g.\ supplied by an external
reference). The cell-type-specific mean expression vector
$\bm{\mu}_j \in \R_{\geq 0}^K$ is identifiable from spot-level data
$\{X_{sj}\}_{s=1\ldots S}$ if and only if $P$ has full column rank
$K$.
\end{theorem}

\begin{proof}
Identifiability of $\bm{\mu}_j$ from the joint distribution of
$\{X_{sj}\}_s$ is equivalent to identifiability of the spot-level
mean vector $\bm{m}_j = P\bm{\mu}_j$, because the Poisson family
is parametrized by its mean (so two parameter vectors yielding the
same mean vector yield the same joint distribution).

\emph{Necessary direction.} Suppose $P$ has column rank less than
$K$, so there exists $\bm{v} \neq \bm{0}$ with $P\bm{v} = \bm{0}$.
For any candidate parameter $\bm{\mu}_j$, the perturbed
$\bm{\mu}_j' = \bm{\mu}_j + \bm{v}$ (chosen so $\bm{\mu}_j'$ remains
in $\R_{\geq 0}^K$, possible whenever $\bm{\mu}_j$ has all positive
entries) satisfies $P\bm{\mu}_j' = P\bm{\mu}_j = \bm{m}_j$. Both
parameters induce identical spot-level distributions of $\{X_{sj}\}$,
so $\bm{\mu}_j$ is not identifiable.

\emph{Sufficient direction.} Suppose $P$ has full column rank $K$.
Then $P^\top P$ is invertible, and the map $\bm{\mu}_j \mapsto
\bm{m}_j = P\bm{\mu}_j$ is injective on $\R^K$: distinct parameters
yield distinct mean vectors, and identifiability of $\bm{\mu}_j$
follows. The closed-form left inverse
$\bm{\mu}_j = (P^\top P)^{-1} P^\top \bm{m}_j$ recovers the
parameter from the mean exactly.
\end{proof}

\begin{remark}[Extensions]
\label{rmk:rank-extensions}
Three extensions of \cref{thm:rank} are immediate. (i) If $N_s$
varies across spots, the proof is unchanged: identifiability is a
property of the parametric family, not the sample size; consistent
estimation of $\bm{m}_j$ requires $\min_s N_s$ or the harmonic mean
to grow, which is the standard Poisson-MLE regularity condition.
(ii) Bernoulli dropout, in which $X_{sj}$ is observed as zero with
some cell-type-independent probability $\pi$, is absorbed into the
mean: $\E[X_{sj}/N_s] = (1-\pi) \sum_k p_{s,k} \mu_{j,k}$, and the
rank condition on $P$ remains necessary and sufficient (up to a
known constant). (iii) Cell-type-dependent dropout $\pi_k$
violates condition C2 of the masking framework and is the regime
treated by \cite{towell2026mdrelax}; the rank condition is no
longer sufficient and identifiability degrades quantitatively in
the gap between the maximum and minimum $\pi_k$.
\end{remark}

\Cref{thm:rank} is the ST analogue of \cref{thm:bg-id}, and it is the
coarsening-framework view of a condition already known in the
deconvolution and matrix-factorization literatures rather than a new
result. With the composition matrix $P$ supplied by a reference, the
full-column-rank requirement is exactly the signature-matrix
identifiability condition behind asymptotic inference for cell-type
proportions \citep{cai2024inference}; the contribution here is to label
it as the coarsening-rank condition shared across the masked-data
series, not to establish it afresh. When $P$ is instead estimated
jointly with $\bm{\mu}_j$ (reference-free deconvolution, e.g.\
STdeconvolve \citep{miller2022reference}), identifiability becomes the
question of when a nonnegative factorization $\bar X \approx P U^\top$
is unique, which the nonnegative-matrix-factorization literature
characterizes through separability and anchor conditions
\citep{donoho2003nmf,arora2012topic,fu2019nmf}; our rank condition is the
reference-supplied special case in which the separability question
collapses to column rank of a known $P$. Read as a mixture-demixing
problem, recovering $\bm{\mu}_j$ from the convex combinations $P\bm{\mu}_j$
is the cell-type instance of demixing mutually contaminated
distributions \citep{blanchard2014decontamination}; we phrase the
solvability condition in the coarsening vocabulary (candidate-set rank)
rather than the contamination-model vocabulary used there.

The same algebra predates spatial deconvolution in the bulk RNA-seq
setting, where a single mixed-tissue measurement is the signature-matrix
product of cell-type proportions and cell-type-specific profiles.
Reference-based bulk deconvolution methods such as MuSiC
\citep{wang2019music}, CIBERSORTx \citep{newman2019digital}, and the
Bayesian BayesPrism \citep{chu2022bayesprism} fit exactly this
signature-times-proportions model, so the full-column-rank requirement of
\cref{thm:rank} and the cell-total identity of \cref{thm:cell-total-st}
apply to them verbatim, with each bulk sample playing the role of a spot
and a cohort of samples playing the role of the spot ensemble. The same
rank deficiency that obstructs recovery there (a signature matrix without
full column rank, or a sample ensemble whose composition matrix is rank
deficient) is the obstruction here. What is genuinely spatial, and absent
in the bulk case, is that the spots come pre-supplied with distinct
compositions $\bm{p}_s$ from tissue architecture and that identifiability
can be restored by single-cell-resolution platforms contributing
unit-vector rows; these are the spatial-specific contributions the
coarsening reading isolates.

\Cref{thm:rank} also clarifies the role of single-cell-resolution
platforms: a single cell measured at single-cell resolution
contributes a row of $P$ that is a unit vector $\bm{e}_k$ (composition
$100\%$ type $k$), maximally informative for column $k$ of $P$. The
classical reliability statement ``singletons restore identifiability''
becomes ``single-cell-resolution spots restore identifiability''.

\subsection{When does the rank condition fail?}

In purely sequencing-based ST (no MERFISH/seqFISH+ integration), $P$
typically has full column rank $K$ \emph{across spots} as long as
spots cover the full cellular diversity of the tissue. The condition
fails when:
\begin{itemize}
\item Two cell types always co-occur in the same proportion (e.g.,
  endothelial and pericyte cells in capillary regions), making their
  contributions inseparable in $P$.
\item The number of distinct cell types $K$ exceeds the number of
  distinct $\bm{p}_s$ vectors observed (rare in practice for
  whole-tissue assays).
\item Cell types are functionally identical with respect to the
  measured genes; in this case the framework is correctly reporting
  that no method can distinguish them.
\end{itemize}

\subsection{Cell-total consistency}

\begin{theorem}[Cell-total consistency]
\label{thm:cell-total-st}
Let $(\hat{\bm{\mu}}_j, \hat{P})$ be an MLE under the spot-level
Poisson likelihood with mean $N_s \sum_k p_{s,k}\mu_{j,k}$, in which
both $\bm{\mu}_j$ and $\bm{p}_s$ are unconstrained at the optimum
(no prior, no penalty, no spatial-smoothness regularization that
ties the two together). Write the residual matrix
$R_{sj} = \bar{X}_{sj} - \sum_k \hat P_{sk}\hat\mu_{j,k}$, where
$\bar{X}_{sj} = X_{sj}/N_s$, and let $\hat U = [\hat\mu_{j,k}] \in
\R^{J \times K}$. At any interior MLE:
\begin{enumerate}
\item[\emph{(a)}] \textbf{Aggregate identity (unconditional).} The
  fitted spot totals match the observed spot totals along the column
  spaces of $\hat P$ and $\hat U$: $\hat P^\top R = 0$ and
  $R\hat U = 0$. In particular, for every spot $s$ the
  composition-weighted gene aggregate is reproduced,
  $\sum_j \hat\mu_{j,k}\,R_{sj} = 0$ for each $k$, and dually
  $\sum_s \hat P_{sk}\,R_{sj} = 0$.
\item[\emph{(b)}] \textbf{Per-entry identity (saturated case).} If in
  addition $\operatorname{rank}(\bar X) \le K$ (the saturated regime,
  in which the observed mean matrix is itself explained by $K$ cell
  types), then $R = 0$ and
  \begin{equation}
    \sum_k \hat{P}_{sk} \hat{\mu}_{j,k} = \bar{X}_{sj}
    \quad\text{for every } (s,j).
    \label{eq:cell-total-st}
  \end{equation}
\end{enumerate}
\end{theorem}

\begin{proof}[Sketch]
The score equation for $\hat\mu_{j,k}$ at an interior optimum reads
$\sum_s \hat P_{sk} R_{sj} = 0$ for each $(j,k)$, i.e. $\hat P^\top R
= 0$; the score equation for $\hat P_{sk}$ reads $\sum_j
\hat\mu_{j,k} R_{sj} = 0$ for each $(s,k)$, i.e. $R\hat U = 0$. This
is part~(a), and it holds unconditionally at any interior MLE.

For part~(b), the two conditions say $R$ lies in the kernel of the
linear operator $\Lambda: R \mapsto (\hat P^\top R,\, R\hat U)$ on
$\R^{S\times J}$. Contrary to a tempting argument, joint full column
rank of $\hat P$ and $\hat U$ does \emph{not} make $\Lambda$
injective: its kernel has dimension $(S-K)(J-K)$, which is positive
whenever $K < \min(S,J)$ (a $2\times 2$, $K=1$ instance has a
one-dimensional kernel, and we confirm the formula numerically in
\cref{sec:validation}). Hence $\hat P^\top R = R\hat U = 0$ does not
force $R = 0$ in general. The per-entry identity holds exactly when
the residual is squeezed out by a rank constraint: if
$\operatorname{rank}(\bar X)\le K$, the fitted $\hat P\hat U^\top$
can match $\bar X$ entrywise at the optimum, so $R = 0$. Otherwise
the unexplained component of $\bar X$ orthogonal to both column
spaces survives, and only the aggregate identity~(a) holds.
\end{proof}

\textbf{Scope and practical implication.} \Cref{thm:cell-total-st}
applies to unregularized MLE-based deconvolution: this covers naive
RCTD and the unregularized limit of joint NMF. It does \emph{not}
cover Cell2location's Bayesian-hierarchical prior or CARD's
spatial-smoothness penalty, which couple $\bm{p}_s$ across spots and
break the unconstrained-optimum assumption. The implication for
unregularized methods is sharp: disagreement on per-cell-type
estimates is invisible in marginal-fit checks. This is the mechanism
behind the empirical finding of \citet{ivich2025missing} that a cell
type missing from the single-cell reference still leaves a well-fitting
aggregate while biasing the per-type estimates, and is detectable only
in the fit residual; their residual-recovery diagnostic is the
empirical counterpart of part~(b) of \cref{thm:cell-total-st}, whereas
we give the exact algebraic conditions (the kernel of $\Lambda$) under
which the per-entry residual vanishes versus persists. For regularized
methods, the closer the regularization is to identity, the more
\eqref{eq:cell-total-st} continues to hold approximately. Diagnosing
per-cell-type bias requires single-cell-resolution data, marker-gene
calibration, or external biological priors. Cell-total consistency
(\cref{thm:cell-total-st}) is a special case of the general
coarsened-data consistency theorem of \citet{towell2026synthesis}, in
which the singleton-candidate-set device used here, single-cell-resolution
probes (MERFISH/seqFISH+), appears as one instance of a
domain-independent rank condition and singleton-restoration mechanism.
